Formalization of Forcing in Isabelle/ZF \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2020-04-16 IJCAR2020/related.tex
Minor corrections and phrasing.

There are various formalizations of Zermelo-Fraenkel set theory in
proof assistants (v.g.\ Mizar, Metamath, and recently Lean
\cite{DBLP:conf/cade/MouraKADR15}) that proceed to different levels of
sophistication. Isabelle/ZF can be regarded as a notational variant of
NGB set theory \cite[Sect.~II.10]{kunen2011set}, because the schemes
of Replacement and Separation feature higher order (free) variables
playing the role of formula variables. It can't be proved that the
axioms thus written correspond to first order sentences. This is the
reason that our relativized versions only apply to set models, where
we restrict those variables to predicates that actually come
from first order formulas. In that sense, the axioms of the locale
\isatt{M{\isacharunderscore}ZF} correspond more faithfully to the
$\ZF$ axioms.

Traditional expositions of the method of forcing
\cite{kunen2011set,Jech_Millennium} are preceded by a study of
relativization and absoluteness. For this reason, it was a natural
choice at the beginning of this project to build on top of Paulson's
formalization of constructibility on Isabelle/ZF, and that was a one
of the main early reasons to work on that logic instead of, e.g., HOL
---there are other reasons, that are discussed below. In any case, our
development of forcing does not depend on constructibility
itself (in contrast to Cohen's
original presentation, in which ground models are initial segments of the
constructible hierarchy).

@@

Concerning the formalization of the method 
of forcing, all of the previous works (to the best of our knowledge)
have been done interpreting set theory into type theory, and
none of them uses the ctm approach. The
most important is the recent \emph{Flypitch} project  by 
Han and van Doorn
\cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes
a formalization of the independence of $\CH$ by means
the Boolean-valued approach to forcing, using the Lean
proof assistant. The Flypitch
formalization is largely orthogonal to ours, and this
precludes a direct comparison of code. But we can
highlight some conceptual differences between our development and the
corresponding fraction of Flypitch.

A first observation concerns consistency strength. The consistency of
Lean requires infinitely many inaccessibles. More precisely, let
Lean$_n$ be the theory of CiC foundations of Lean restricted to $n$
type universes.  Carneiro \cite{carneiro-ms-thesis}, proved the
consistency of Lean$_n$ from $\ZFC$ plus the existence of $n$
inaccessible cardinals. It is also reported in
\cite{carneiro-ms-thesis} that Werner's results in
\cite{10.5555/645869.668660} can be adapted to show that Lean$_{n+2}$
proves the consistency of the latter theory.  In that sense, although
Flypitch includes proofs of unprovability results in first order
logic, the meta-theoretic machinery used to obtain them is far heavier
than the one we use to operate model-theoretically.

>>>>

There are various formalizations of Zermelo-Fraenkel set theory in
proof assistants (v.g.\ Mizar, Metamath, and recently Lean
\cite{DBLP:conf/cade/MouraKADR15}) that proceed to different levels of
sophistication. Isabelle/ZF can be regarded as a notational variant of
NGB set theory \cite[Sect.~II.10]{kunen2011set}, because the schemes
of Replacement and Separation feature higher order (free) variables
playing the role of formula variables. It cannot be proved that the
axioms thus written correspond to first order sentences. For this
reason, our relativized versions only apply to set models, where
we restrict those variables to predicates that actually come
from first order formulas. In that sense, the axioms of the locale
\isatt{M{\isacharunderscore}ZF} correspond more faithfully to the
$\ZF$ axioms.

Traditional expositions of the method of forcing
\cite{kunen2011set,Jech_Millennium} are preceded by a study of
relativization and absoluteness. For this reason, it was a natural
choice at the beginning of this project to build on top of Paulson's
formalization of constructibility on Isabelle/ZF, and that was one
of the main early reasons to work on that logic instead of, e.g., HOL
---below we discuss other reasons. In any case, our
development of forcing does not depend on constructibility
itself (in contrast to Cohen's
original presentation, in which ground models are initial segments of the
constructible hierarchy).

@@

Concerning the formalization of the method 
of forcing, all of the previous works (to the best of our knowledge)
have been done interpreting set theory into type theory, and
none of them uses the ctm approach. The
most important is the recent \emph{Flypitch} project  by 
Han and van Doorn
\cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes
a formalization of the independence of $\CH$ by means of
the Boolean-valued approach to forcing, using the Lean
proof assistant. The Flypitch
formalization is largely orthogonal to ours, and this
precludes a direct comparison of code. But we can
highlight some conceptual differences between our development and the
corresponding fraction of Flypitch.

A first observation concerns consistency strength. The consistency of
Lean requires infinitely many inaccessibles. More precisely, let
Lean$_n$ be the theory of CiC foundations of Lean restricted to $n$
type universes.  Carneiro proved in his MSc thesis~\cite{carneiro-ms-thesis} the consistency of Lean$_n$ from $\ZFC$ plus
the existence of $n$ inaccessible cardinals. It is also reported in
Carneiro's thesis that Werner's results in
\cite{10.5555/645869.668660} can be adapted to show that Lean$_{n+2}$
proves the consistency of the latter theory.  In that sense, although
Flypitch includes proofs of unprovability results in first order
logic, the meta-theoretic machinery used to obtain them is far heavier
than the one we use to operate model-theoretically.
